---
title: "Analyzing the Impact of Location and Time on High-Value Bicycle Thefts in Toronto."
subtitle: "Revealing Seasonal Trends, Hotspots, and Predictive Factors."
author: "Ziyi Liu"
date: "`r Sys.time()`"
date-format: "D MMMM YYYY"
abstract: "Bicycle theft is one of the major problems that cities have to deal with, as it not only affects individuals but also communities. The complex nature of bicycle theft patterns and risk factors is at the core of creating preventive measures that actually work. In the project, we will examine the bicycle theft data from the Toronto Police Service in the temporal trends, spatial patterns, and the social-economic factors that influence the incidences of thefts. Using statistical techniques like logistic regression and machine learning algorithms, we determine the key predictors of theft chance: theft location, bike characteristics, and temporal aspects. We are able to identify the specific seasonal fluctuations and the hourly patterns in which the thefts are most likely to happen, thus emphasizing the need for tailored prevention measures. Besides, we examine the sociocultural and environmental influences leading to thefts, giving ideas for integrated prevention approaches. The project is designed to combine a quantitative analysis with a qualitative insight with the purpose of informing evidence-based policies and interventions that would reduce the level of bicycle theft in the city and enhance urban security."
format: 
  pdf:
    documentclass: article
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

# SET UP WORKSPACE
#install.packages("tidyverse")
#install.packages("knitr")
#install.packages("here")
#install.packages("modelsummary")
#install.packages("gridExtra")
#install.packages("car")
#install.packages("pROC")
library(tidyverse)
library(knitr)
library(here)
library(modelsummary)
library(gridExtra)
library(car)
library(pROC)
```

```{r}
#| include: false

# IMPORT PROCESSED DATA FROM OUTPUTS FOLDER 
# imported via 00-data_cleaning.R file

data <- read.csv(here::here("outputs/data/cleaned_data.csv"))

```

# Introduction

This study is about urban bicycle theft and attempts to illuminate the different chances of high-priced bicycles being shoplifted in public or private places. This study is unique in the sense that it looks at multiple factors and thus takes into account the time of day, brand, the color of the bike, and seasonal variations. This goal is not only defined by dedication to community safety and the promotion of biking but also by the need to find out unknown contexts and patterns of bicycle theft that have not been researched before [@mburu2016environmental].

This study’s central hypothesis is the fact that the interaction of the district where the theft occurs and such nuanced variables play a crucial role in the outcomes of the crime. This research is different from the previous studies, which either looked at theft incidents broadly or targeted only singular aspects like the effectiveness of security interventions or the effect of environmental factors. Instead, this study dives into the minutiae of how and why certain bicycles get stolen particularly in the context of specific settings [@levy2018case]. A mixed-method study that will involve both quantitative theft data as well as qualitative information from cyclists and law enforcement will be used in order to offer a broad perspective of the dynamics of bicycle theft.

The variables that are chosen – which are temporal factors like time and season and the intrinsic features of the bikes such as brand popularity and color – are inspired by a gap identified in the literature [@salvanelli2019predicting]. Such decisions are made with the intention of increasing the scope of the dataset by focusing only on urban theft cases thus offering a richer and more detailed study than was the case before. The outcomes of this research seek not only to add to the academic discourse by identifying new patterns and preventive measures but also to offer practical recommendations for urban planning and community efforts aimed at minimiz- ing theft-related losses. 

By filling this critical research void, the findings of this study are anticipated to inform targeted interventions, enriching the strategies employed by urban planners and cycling advocates to safeguard bicycles. In doing so, this research stands as a pioneering effort to explore the multifaceted nature of urban bicycle theft, providing a solid foundation for future studies while contributing actionable insights toward the creation of safer, more bike-friendly urban environments.

The structure of this paper is organized as follows: @sec-data introduces the data used for analysis, including visualizations of the variables of interest; @sec-model presents the logistic regression model combined with lasso used to explore the relationships between bicycle theft variables; @sec-results displays the interpretations of the model alongside other findings; and @sec-discussion provides a discussion on the implications of the findings, the limitations of this study, and its next steps for further research on this topic. 

Programming scripts and data are available at: https://github.com/liuziy23/torontobiketheftanalysis.

# Data {#sec-data}

## Data Source {#sec-variables-of-interest}

The data utilized in this study focuses on urban bicycle thefts, accessible through the open database provided by the Toronto Police Service, available at: Toronto Police Service - Bicycle Thefts [@citebiketheft]. The database contains detailed records of reported bicycle thefts in Toronto, covering various aspects including the time, location, type of premises, bike make, and color, among other factors. The open-source nature of this dataset allows for free public access, supporting research and community awareness efforts regarding bicycle theft.

The primary entities of this dataset include the theft incident, which is characterized by specific details such as the type of bicycle, the circumstances of the theft, and the outcome if known. Additional elements captured in the dataset include the make and color of the bicycle, the location of the theft, and the date and time it occurred, along with identifiers for each unique incident.

The procedure of data collection includes reports from the victims of theft, the police and the eyewitnesses. Data collection is based on incident reporting which, in distinction to convenience sampling, does not possibly cover all demographic and geographic areas but provides real, specific crime case stories. This method is very useful as it offers the most authentic cases and events of the bikes being stolen, providing a practical look at the patterns and preventive methods.

The reports are used to study the deeper issues of urban bike theft such as the effectiveness of bike security measures and role of the environment in bike thefts. These reports represent the comprehensive picture of bicycle theft that occur throughout the year, illustrating the ongoing challenge of keeping bicycles secure in a crowded place. The data set is enhanced with the location and time of thefts that are crucial for the analysis of crime patterns and thinking about ways to solve the problem in the community.

This analysis is performed in R language which is a programming language [@citeR] and is used for statistical computations and visualizing data. Essential packages from the tidyverse [@citetidyverse] are employed, including dplyr [@citedplyr] for data manipulation and cleaning, readr [@citereadr] for data importation, here [@citehere] for specifying file paths, ggplot2 [@citeggplot2] for generating data visualizations, gridExtra [@citegridExtra] for arranging visualizations, knitr [@citeknitr] for formatting output tables, and modelsummary [@citemodelsummary] for producing summary tables. Additionally, the glmnet package [@citeglmnet] is used for regularization techniques and model fitting, the car package [@citecar] provides advanced data handling and statistical tests to including assessing multicollinearity, and the pROC package [@citepROC] provides tools for calculating measures from logistic regression models.

## Variables of Interest

Our analysis specifically focuses on the incidence of bicycle thefts. In this paper, we utilize a dataset exclusively dedicated to bike thefts, which contains 32,874 observations, where each row corresponds to a reported theft incident and each column represents a distinct variable, such as demographic details or the circumstances surrounding the theft [@citebiketheft]. This dataset enables us to explore a variety of factors including the type of location where the theft occurred, the brand of the bicycle, the time and month of the theft, the color of the bicycle, and the value of the bicycle. It is noteworthy that the 32,874 records in our dataset cover a wide range of theft incidents, allowing for a comprehensive analysis of trends and patterns.

Variables included in the analysis are:

- Location Type: Main exposure variable, distinguishes theft location as public or private; included as a predictor.

- Bike Make: Brand of bicycle, used to explore brand's impact on theft likelihood; included as a predictor.

- Occurrence Hour: Time of theft, used to identify high-risk hours; included as a predictor.

- Occurrence Month: Month of theft, used to analyze seasonal theft patterns; included as a predictor.

- Bike Colour: Assesses if color influences theft risk; included as a predictor.

- Bike Value: Binary response variable; categorized as high value (over $1000) or not.

\newpage

```{r}
#| label: fig-1
#| fig-cap: Distribution of location type, hour of the day, bike colour, and bike value.
#| echo: false
#| warning: false

# Individual plots for each variable
p1 <- ggplot(data, aes(x = location)) +
  geom_bar() +
  labs(title = "Figure 1.", 
       subtitle = "Distribution of Theft Locations", x = "Location Type", y = "Count") + theme_minimal()

p2 <- ggplot(data, aes(x = BIKE_MAKE)) +
  geom_bar() +
  labs(title = "Figure 2.", 
       subtitle = "Distribution of Bike Makes", x = "Bike Make", y = "Count") +  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

p3 <- ggplot(data, aes(x = OCC_HOUR)) +
  geom_bar() +
  labs(subtitle = "Distribution of Theft Hours", x = "Hour of the Day", y = "Count") + theme_minimal()

p4 <- ggplot(data, aes(x = OCC_MONTH)) +
  geom_bar() +
  labs(subtitle = "Distribution of Theft Months", x = "Month", y = "Count") + theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p5 <- ggplot(data, aes(x = BIKE_COLOUR)) +
  geom_bar() +
  labs(subtitle = "Distribution of Bike Colours", x = "Bike Colour", y = "Count")+ theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p6 <- ggplot(data, aes(x = value)) +
  geom_bar() +
  labs(subtitle = "Distribution of Bike Values", x = "Bike Value", y = "Count") + theme_minimal()

# Combining plots with grid.arrange
grid.arrange(p1, p3, p5, p6, ncol = 2)
```

```{r}
#| label: fig-2
#| fig-cap: Distribution of bike make and theft-occurrence month.
#| echo: false
#| warning: false

grid.arrange(p2, p4, ncol = 2)

```

In this article, we develop the visualizations of the distributions of the variables of interest @fig-1 and @fig-2. These figures are the ones that make it possible to show the different structures and forms that make each variable to have a different distribution. @fig-1 shows the distribution of theft occurrences by day/time of day and month, uncovering certain time and seasonal patterns that might be indicative of higher theft risks at some hours and during specific seasons. For instance, the visualization may show the rise in the thievery at late evenings and in the warmer months so that the security measures should be taken into consideration at that time. On the contrary, @fig-2, shows the distribution of the bikes brand, color, and price. These figures give us perspectives on which bike brands are more often stolen, what colors are more susceptible to theft, and how often high-value bicycles are involved in theft cases. For instance, high reports of bike thefts that concentrate on stealing expensive bikes of specific color implies that bike thieves might have certain preferences in the market.

The visualizations perform two important functions. First, they help to interpret the raw data. Second, they become a basis for the following statistical modeling phase. The varied structures observed in the data distributions—ranging from skewed distributions for bike values to categorical distinctions in bike make and color—will be carefully accounted for in our modeling strategies. These efforts are aimed at developing a comprehensive understanding of the factors influencing bicycle theft, which in turn can inform targeted interventions to reduce such incidents. By meticulously analyzing the distribution and influence of each variable, we enhance the predictive power of our models, thereby contributing valuable insights into effective theft prevention strategies.

\newpage

Understanding when bike thefts occur helps target prevention efforts. Seasonal and hourly trends indicate times of higher theft risk, guiding strategic security planning and resource allocation.

```{r}
#| label: tbl-month
#| tbl-cap: Summary statistics table of occurrence month.
#| echo: false

# Summarize categorical variable 'OCC_MONTH' for frequency and percentage
categorical_summary <- data %>%
  group_by(OCC_MONTH) %>%
  summarise(
    Frequency = n(),
    Percentage = round((n() / nrow(data) * 100),2)
  ) %>%
  ungroup() %>%
  arrange(desc(Frequency))

# Print categorical summaries
kable(categorical_summary)
```

The categorical summary of bike theft occurrences across various months, as presented in @tbl-month The table captures 32,874 instances of reported bike thefts, with each month revealing a distinct pattern of frequency. Notably, July stands out as the month with the highest theft incidents, constituting approximately 15.52% of the total cases. This is closely followed by August and June, with 14.38% and 13.44% respectively, indicating a trend of increased thefts during the summer months. Conversely, the winter months—specifically February and January—show a marked decrease in theft frequency, accounting for only 2.53% and 2.73% of the incidents. Within the table, the 'Frequency' column enumerates the actual count of thefts per month, while the 'Percentage' column elucidates the proportion of total thefts that occurred in each month. This distribution highlights a clear seasonal pattern, with more than 40 percent of thefts taking place during the warmer months of July through September, and the fewest thefts occurring in the cold of December through February.

\newpage

```{r}
#| label: tbl-hour
#| tbl-cap: Summary statistics table of occurrence hour
#| echo: false

# Summarize categorical variable 'OCC_HOUR' for frequency and percentage
categorical_summary_hour <- data %>%
  group_by(OCC_HOUR) %>%
  summarise(
    Frequency = n(),
    Percentage = round((n() / nrow(data) * 100), 2)
  ) %>%
  ungroup() %>%
  arrange(desc(Frequency))

# Print categorical summaries
kable(categorical_summary_hour)
```

The comprehensive summary of bike theft occurrences delineated by hour, as depicted in @tbl-hour, outlines the frequency and percentage of thefts at various times throughout a 24-hour period, based on 32,874 reported incidents. The hour with the highest frequency of thefts is 18:00 (or 6 PM), representing 7.05% of total incidents, suggesting an elevated risk of theft during early evening hours. This is followed by hours 17:00 (6.49%) and 12:00 (6.01%), indicative of heightened theft activity in late afternoon and midday, respectively. In stark contrast, the early hours of the morning—specifically 01:00 and 05:00—show significantly fewer occurrences, with the lowest being at 5 AM, constituting only 0.11% of thefts. The 'Frequency' column in the table provides the precise count of thefts at each hour, while the 'Percentage' column offers insight into the share of total thefts happening at each corresponding hour. The data elucidates a pronounced pattern of theft risk that peaks during evening hours and diminishes substantially in the early morning, demonstrating the temporal trends in bicycle theft incidents.

# Model {#sec-model}

```{r}
#| include: false

# IMPORT MODEL FROM OUTPUTS FOLDER
logit_model <- readRDS(here::here("outputs/models/logit_model.rds"))

```

## Model Methods - Variable Selection

In our study to discern the determinants of high-value bicycle theft, we utilize a methodological framework that combines Lasso logistic regression with cross-validation. Logistic regression is a very appropriate method for our task because it is especially good at modeling binary outcomes, a situation similar to classifying incidents of high-value bicycle theft. Its capacity to produce probability figures together with odds ratios for the occurrence of an event, depending on different factors, makes it a critical tool in uncovering the likelihood of high-end thefts. This method enables us to translate the features of the bikes and their setting into the probability of the bikes being targeted by thieves, giving results that are both statistically significant and practical.

The use of Lasso (Least Absolute Shrinkage and Selection Operator) in the logistic regression model makes the selection of variables possible, especially in cases where there is an excess of predictors. The main issue in urban crime data is their high dimensionality. Thus, many variables like time, location, brand of a bike, and seasonality have a chance to impact theft risk. Lasso manages this complexity by effecting the coefficient of the unimportant variables which are subsequently shrunk to zero. This not only allows finding the most prominent predictors but also suppresses the risk of overfitting, thus increasing the model's predictive value and clarity.

From the logistic regression calculations based on Lasso variable selection, conclusions will be made, where the variables and their odds ratio will be the main focus. These odds ratios are being used to assess the strength and the direction of the relationships between each predictor and the probability of a bicycle being stolen as a high value. The statistical significance of these associations along with applied factors determine how we perceive the dynamics of high-value bike theft incidents.


## Model Methods - Model Diagnostics and Validation

Model diagnostics includes the evaluation of Variance Inflation Factor (VIF) and Receiver Operating Characteristic (ROC) curve with the area under curve (AUC) metric. Post-Lasso variable selection, VIF is used to detect multicollinearity among predictors. A VIF above 5 indicates problematic multicollinearity, potentially distorting the model's coefficients. Addressing this might involve dropping correlated predictors to improve model stability and interpretability. The ROC curve, plotted by comparing true positive rates against false positive rates at various thresholds, alongside the AUC metric (area under the ROC curve), evaluates the model's ability to distinguish between binary outcomes effectively. An AUC close to 1 signifies excellent model performance, while values near 0.5 suggest no predictive benefit over random chance. An AUC over 0.6 will be deemed as adequately well model performance. These diagnostics ensure our model's robustness by mitigating multicollinearity and verifying its predictive accuracy.

Model validation is conducted via cross-validation. Cross-validation is integral to our methodology, serving the dual purpose of validation and optimization. By partitioning the data into numerous sets and iteratively using each set as a validation while training on the others, we ascertain the model's performance across different subsets of data. This process is key to determining the optimal lambda $\lambda$, the regularization parameter in Lasso that influences the degree of shrinkage applied to the coefficients. The optimal lambda is the one that minimizes the cross-validated error, striking a balance between the model's complexity and its ability to generalize to new data.

\newpage

# Results {#sec-results}

```{r}
#| label: tbl-modelsum
#| tbl-cap: Model summary of the fitted logistic regression model (Model 1) on bicycle theft being of high value.
#| echo: false

kable(summary(logit_model)$coefficients[,c(1,2,4)], digits=2)
```

The final model, referred as Model 1, is then a logistic regression model with bike theft being high value or not as the response variable and predictors being bike make, bike color, and location of theft. @tbl-modelsum showcases the coefficients of the predictor variables of Model 1. We focus on the upper section of this table, which includes the intercept coefficient and the coefficient representing the log odds of a stolen bike being of high value, and examines how they vary with different predictors. The standard errors of the estimated regression coefficients are displayed in brackets for reference.

We see location of theft, make of bike and colour of bike are statistically significant at 5%. Significant coefficients greater than 0 indicates being associated with more likely to be stolen while less than 0 is associated with less likely to be stolen. Bikes parked outside are less likely to be stolen. All bike makes are less likely to be stolen than bike make of "CA", except for "specialized" makes which are more likely to be stolen than "CA" bikes. All bike colours are less likely to be stolen than black bikes, except for white bikes which are equally likely to be stolen than black bikes.

```{r}
#| label: fig-roc
#| fig-cap: ROC and AUC of the fitted logistic regression model on bicycle theft being of high value.
#| echo: false
#| warning: false

library(pROC)
predicted_probs <- predict(logit_model, type = "response")
roc_obj <- roc(data$value, predicted_probs)
plot(roc_obj, main = "Figure 3. ROC Curve", col = "#1c61b6", lwd = 2)
abline(a = 0, b = 1, lty = 2, col = "gray")
text(0.6, 0.2, paste("AUC =", round(auc(roc_obj), 3)), col = "#1c61b6")
```

@fig-roc shows the ROC curve and the AUC. The AUC is 0.647, which means the model has adequate prediction accuracy. 

```{r}
#| label: tbl-vif
#| tbl-cap: VIFs of the fitted logistic regression model (Model 1) on bicycle theft being of high value.
#| echo: false

b <- as.data.frame(vif(logit_model))
b <- data.frame(Predictor = rownames(b),
                VIF = as.numeric(b[,1]))
kable(b, digits=2)
```

The predictors in the model all have a VIF less than 2 and being very close to 1, as shown in @tbl-vif, indicating lack of multicollinearity in the model. 

\newpage

# Discussion {#sec-discussion}

## Understanding the complexities of bicycle theft reveals insights into theft patterns and risk factors {#sec-first-point}

The final logistic regression model investigates the influence of different factors on bicycle theft likelihood, and value (i.e., high-value or not) is the focus. Importantly, the model exemplify the importance of the theft location, where bikes parked outside comes in with a coefficient less than 0. This implies that the bicycles parked outside are not likely to be classified as high- value thefts like those parked in unspecified default locations where the possibility of being more secure or private areas is higher. In addition, an interesting observation regarding bike model is noted showing that all models are generally less likely to be stolen except for "CA" bikes except for the "Specialized" brand which stands out with a positive coeﬀicient. This means that ” Specialized” bikes which are considered to be of good quality and might have better resale value, are more vulnerable to theft, thus showing the importance of brand perception for theft risk. In addition, with regard to the color of bikes, all colors except white are less frequently linked to expensive bike thefts, whereas black bikes are. The equally high chances of white bikes being stolen as black bikes may be a reflection of a specific preference or market demand among criminals for these colors, possibly because of their visibility or sales value. Finally, the model reveals the complicated interaction between theft location, bike manufacturer, and bike color in determining the probability of a bicycle being a high-value theatrical target, thus highlighting the critical role of external factors in crime likelihood. This analysis directly answers our research question by pointing out certain characteristics of a bicycle that makes it more vulnerable to theft; thus, I will suggest targeted counter-measures to combat high-value bicycle thefts.

## Investigating the temporal trends of bicycle thefts provides actionable insights for theft prevention efforts {#sec-second-point}

From the analysis of the monthly and hourly bicycle theft occurrence, the temporal patterns are observed, which are the cornerstone of strategic theft prevention measures. The study reveals a well-defined seasonal pattern, in which there is more theft in the summer and less in the winter. This seasonal change reveals the necessity of adjusting the theft prevention strategies depending on the seasons with more crime activities, for example, intensifying patrols and security measures during the peak theft seasons. Moreover, this type of crime mapping helps to show when thieves are more likely to be active and allows to target surveillance and intervention efforts. The data on these temporal patterns can be used by decision-makers to apply preventive measures in time and ensure the safety of bike owners and urban areas.

## Understanding the socio-economic and environmental factors contributing to bicycle thefts informs holistic prevention strategies {#sec-third-point}

Bike theft is a multi-faceted issue that is impacted by a range of socio-economic and environmental factors. Consequently, an integrated approach is essential to stop this kind of theft. The factors such as income inequality, secure parking facilities and urban infrastructure are among the fundamental causes of robbery. Targeted interventions, that deal with the root causes, like providing an efficient bike parking infrastructure and community policing in high theft areas, should be implemented to address theft risk and also contribute to the security of the urban areas. It is also possible through the public awareness campaigns and community involvement initiatives the people can be given the sense of shared responsibility and awareness that will lead to their active participation in theft prevention measures. Through solving the problem of bicycle thefts and joint efforts, communities will be able to turn the cycling experience into safe one and, as a result, reduce the number of bike thefts.

## Limitations & Next Steps {#sec-limitations-next-steps}

### Limitations {#sec-limitations}

The final logistic regression model, although it may provide valuable insights into what determines high-value bicycle theft, is not without its limitations. One of the problems is the AUC of the ROC curve which is 0.647. This metric, indicating that the model can distinguish between high-value and low-value thefts with some degree of accuracy, can be interpreted as moderate predictive accuracy. The AUC value being much nearer to 0.5 than one confirms that the model’s ability to precisely classify high-value thefts based on the predictors-bike make, bike color, and location of theft is very limited. This moderate AUC value may negatively impact the applicability of the model in the real world. For stakeholders, such as urban planners and law enforcement agencies, who intend to use this model to deploy targeted interventions, there is a risk that the strategies might not be as effective as anticipated in preventing valuable bicycle thefts. The model’s insufficient discriminative power could cause misclassification of theft risks, which could lead to inappropriate use of resources and the sub-optimal design of prevention measures.

A number of constraints are behind the AUC restriction that could not be eliminated within the framework of this study. Firstly, the complexity of human behavior, particularly in the case of theft, is hard to adequately portray with bike make, color, and location data only. The non-accounting of such variables as security devices or surveillance in the model could be the reason for the degree of theft likelihood being influenced significantly. The second factor is that data quality and completeness can be critical. With missing or incorrectly reported theft incidences, the relationship between the predictors and the response may be wrong, thus affecting the model’s reliability.

### Next Steps {#sec-next-steps}

Though the model has some limitations in representing the dynamics of high-value bicycle theft, it still gives us an introductory level of understanding. It underlines the factors of major risk that can be used as grounds for initial guidelines for the prevention of theft. Future investigations could seek to improve model accuracy by the addition of predictors, such as time of the day or security measures, and the usage of more complicated modeling methods that can deal with the intricacy of predicting theft events. The continuous process aided by the existing model’s insights will help us to develop our knowledge of the phenomenon and come up with ways of preventing high-value bicycle thefts.

\newpage

# References {#sec-references}
